A Decision Procedure for (Co)datatypes in SMT Solvers

### 2016-03-02 jour/jour.tex
minor changes
added URL

\newcommand\xend{{\hfill$\scriptstyle\blacksquare$}}
\newcommand\xqed{{\hfill$\scriptstyle\squareforqed$}}

@@

\paragraph{Related Work.} Barrett
et al.\ \cite{barrett-et-al-2007} provide a good account of
related work on datatypes as of 2007, in addition to describing their
implementation in CVC3.
Since then, datatypes have been added not only to CVC4 (a complete rewrite of CVC3)
but also to the SMT solver
Z3 \cite{de-moura-bjoerner-2008} in unpublished work by Leonardo de Moura and to a SPASS-like %prototypical
superposition prover by Wand \cite{wand-2014}.
In his Ph.D.\ thesis \cite{bjoerner-1998}, Bj\o rner introduced a decision procedure for
(co)datatypes in STeP, the Stanford Temporal Prover.
Closely related are the automatic structural
% cruanes-201x,wand-weidenbach-201x,  -- both were rejected
induction in SMT solvers
\cite{reynolds-kuncak-2014}
and superposition provers\cite{kersani-peltier-2013,cruanes-2015},
the (co)datatype and (co)induction support in Dafny \cite{leino-moskal-2014},
and the (\hbox{semi-})decision procedures for datatypes
implemented in Leon \cite{suter-et-al-2011} and RADA \cite{pham-et-al-2013}.
%
Datatypes are supported by the higher-order model finder Refute
\cite{weber-2008} for Isabelle. Its successor, Nitpick
\cite{blanchette-nipkow-2010}, can also generate models involving cyclic
codatatype values. Cyclic values have been studied extensively under the heading
of regular or rational trees---see Carayol and Morvan
\cite{carayol-morvan-2006} and Djellou et al.\ \cite{djelloul-et-al-2008}
for recent work.
The $\mu$-notation is inspired by the $\mu$-calculus
\cite[etc.]{kozen-1983,endrullis-et-al-2011}.

@@

\begin{examplex}
\label{ex:model}
Continuing Example~\ref{ex:assign-complete}, let $\J$ be a normal interpretation, where
up to renaming of $\MU$-bound variables, we have
\[\begin{array}{r@{}c@{}r@{}c@{}l@{\qquad}r@{}c@{}l@{\qquad}r@{}c@{}l}
& &\interp{\J}{u} & {}\,=\,{} &  \const{C}( \MU {\Varec{z}}.\; \const{E}( {\Varec{z}} ) ) &
\interp{\J}{w} & = &  \const{E}( \MU {\Varec{y}}.\; \const{D}( {\Varec{y}} ) ) &
\interp{\J}{y} & {}\,=\,{} & \MU {\Varec{y}}.\; \const{D}( {\Varec{y}} ) \\
\interp{\J}{v} & = & \interp{\J}{s} & = & \const{D}( \MU {\Varec{z}}.\; \const{E}( {\Varec{z}} ) ) &
\interp{\J}{x} & {}\,=\,{} & \const{C}( \const{D}( \MU {\Varec{z}}.\; \const{E}( {\Varec{z}} ) ) ) &
\interp{\J}{z} & = & \MU {\Varec{z}}.\; \const{E}( {\Varec{z}} )
\end{array}\]
It can be shown that $\J$ satisfies $\Ec =
\{
u \teq \const{C}( z ),\;
v \teq \const{D}( z ),\;
w \teq \const{E}( y ),\;
x \teq \const{C}( v ),\;
v \teq s,\;
z \not\teq v
\}.$
\xend
\end{examplex}

@@

\begin{enumerate}
\item Let $U$ be the datatype equivalence classes of $\Ec.$
\item Repeat until $U \neq \emptyset$:
\begin{itemize}
\item[\ ] For some $\ec{t} \in U,$ $\rn{Traverse}( \ec{t}, U, \emptyset ).$
\end{itemize}
\end{enumerate}

@@

\begin{enumerate}
\item If $\ec{t} \in P,$ $\rn{Acyclic}(t).$
\item If $\ec{t} \in U$:
\begin{enumerate}
\item[2.1.] If $\const{C}( t_1, \ldots, t_k ) \in \ec{t},$ then for each $j = 1, \ldots k$:
\begin{itemize}
\item[2.1.1.] $\rn{Traverse}( \ec{t_j}, U, P \cup \{ \ec{t} \} ).$
\end{itemize}
\item[2.2.] $U := U \textbackslash \{ \ec{t} \}.$
\end{enumerate}
\end{enumerate}
\vspace*{-\topsep}
\caption{\,Algorithm for checking \rn{Acyclic}}
\label{fig:acyclic-algo}
\end{figure}

@@

\item Repeat until $\mathcalx{U}$ is unchanged:
\begin{enumerate}
\item[2.1.] $\mathcalx{U}' := \emptyset.$
\item[2.2.] For each $U_i \in \mathcalx{U}$:
\begin{itemize}
\item[2.2.1.] Let $\mathcalx{U}_{\vthinspace i}$ be a partition of $U_i$ such that $\ec{t} =_{\mathcalx{U}_{\vthinspace i}} \ec{u}$ for distinct $\ec{t}, \ec{u}$ if and only if
\begin{itemize}
\item $\const{C}( t_1, \ldots, t_k ) \in \ec{t}$ and $\const{C}( u_1, \ldots, u_k ) \in \ec{u},$ and
\item $\ec{t_j} =_{\mathcalx{U}} \ec{u_j}$ for each $j = 1, \ldots, k.$
\end{itemize}
\item[2.2.2.] $\mathcalx{U}' := \mathcalx{U}' \cup \mathcalx{U}_{\vthinspace i}.$
\end{itemize}
\item[2.3.] $\mathcalx{U} := \mathcalx{U}'.$
\end{enumerate}
\item If distinct $\ec{t}, \ec{u} \in U$ for some $U \in \mathcalx{U},$ $\rn{Unique}(t,u).$
\end{enumerate}
\vspace*{-\topsep}
\caption{\,Algorithm for checking \rn{Unique}}
\label{fig:uniqueness-algo}
\end{figure}

@@

By contrast, the contributions of the decision procedure
to proving are less obvious; they depend on how
often acyclicity and uniqueness are necessary for a proof.
To evaluate this,
we generated benchmark problems from existing interactive
proof goals arising in existing
Isabelle formalizations, using Sledgehammer \cite{blanchette-et-al-2013-smt}
as translator from Isabelle to SMT-LIB.
%There are not yet many benchmarks available, since codatatypes are a
%recent addition to Isabelle \cite{blanchette-et-al-2014-impl}. \rem{Necessary to mention this? Seems to detract from significance of contribution.}
We
included all the formalizations from the Isabelle distribution (Distro, 1179~goals)
and the \emph{Archive of Formal Proofs} (AFP, 3014~goals) \cite{klein-et-al-afp}
that define codatatypes falling within the supported fragment. We
also included formalizations about Bird and Stern--Brocot trees (\gandl,
265~goals) \cite{gammie-lochbihler-2016}. To exercise the datatype support, formalizations about finite
lists and trees
were added to the first two benchmark sets. The formalizations were selected before
conducting any experiments. The experimental data are available
online.%
\footnote{\url{http://lara.epfl.ch/~reynolds/JAR-CADE2015-cdt/}}

>>>>

\newcommand\xend{{\hfill\llap{$\scriptstyle\blacksquare$}}}
\newcommand\xqed{{\hfill$\scriptstyle\squareforqed$}}

@@

\paragraph{Related Work.} Barrett
et al.\ \cite{barrett-et-al-2007} provide a good account of
related work on datatypes as of 2007, in addition to describing their
implementation in CVC3.
Since then, datatypes have been added not only to CVC4 (a complete rewrite of CVC3)
but also to the SMT solver
Z3 \cite{de-moura-bjoerner-2008} in unpublished work by Leonardo de Moura and to a SPASS-like %prototypical
superposition prover by Wand \cite{wand-2014}.
In his Ph.D.\ thesis \cite{bjoerner-1998}, Bj\o rner introduced a decision procedure for
(co)datatypes in STeP, the Stanford Temporal Prover.
Closely related are the automatic structural
% cruanes-201x,wand-weidenbach-201x,  -- both were rejected
induction in SMT solvers
\cite{reynolds-kuncak-2014}
and superposition provers\cite{kersani-peltier-2013,cruanes-2015},
the (co)datatype and (co)induction support in Dafny \cite{leino-moskal-2014},
and the (\hbox{semi-})decision procedures for datatypes
implemented in Leon \cite{suter-et-al-2011} and RADA \cite{pham-et-al-2013}.
%
Datatypes are supported by the higher-order model finder Refute
\cite{weber-2008} for Isabelle. Its successor, Nitpick
\cite{blanchette-nipkow-2010}, can also generate models involving cyclic
codatatype values. Cyclic values have been studied extensively under the heading
of regular or rational trees---see Carayol and Morvan
\cite{carayol-morvan-2006} and Djellou et al.\ \cite{djelloul-et-al-2008}
for recent work.
The $\mu$-notation is inspired by the $\mu$-calculus
\cite{kozen-1983,endrullis-et-al-2011}.
% [etc.]

@@

\begin{examplex}
\label{ex:model}
Continuing Example~\ref{ex:assign-complete}, let $\J$ be a normal interpretation, where
up to renaming of $\MU$-bound variables, we have
\[\begin{array}{r@{}c@{}l@{\qquad}r@{}c@{}l@{\qquad}r@{}c@{}l}
\interp{\J}{u} & {}\,=\,{} &  \const{C}( \MU {\Varec{z}}.\; \const{E}( {\Varec{z}} ) ) &
\interp{\J}{w} & = &  \const{E}( \MU {\Varec{y}}.\; \const{D}( {\Varec{y}} ) ) &
\interp{\J}{y} & {}\,=\,{} & \MU {\Varec{y}}.\; \const{D}( {\Varec{y}} ) \\
\interp{\J}{s} \,=\, \interp{\J}{v} & = & \const{D}( \MU {\Varec{z}}.\; \const{E}( {\Varec{z}} ) ) &
\interp{\J}{x} & {}\,=\,{} & \const{C}( \const{D}( \MU {\Varec{z}}.\; \const{E}( {\Varec{z}} ) ) ) &
\interp{\J}{z} & = & \MU {\Varec{z}}.\; \const{E}( {\Varec{z}} )
\end{array}\]
%It can be shown that
It is easy to see that $\J$ satisfies the constraints

\vskip\abovedisplayskip

\noindent
\hbox{}{\color{white}\xend}$u \teq \const{C}( z )$\hfill
$v \teq \const{D}( z )$\hfill
$w \teq \const{E}( y )$\hfill
$x \teq \const{C}( v )$\hfill
$v \teq\nobreak s$\hfill
$z \not\teq v$\xend
\end{examplex}

@@

\begin{enumerate}
\item Let $U$ be the datatype equivalence classes of $\Ec.$
\item Repeat until $U \neq \emptyset$:
\begin{itemize}
\item[2.1.] For some $\ec{t} \in U,$ $\rn{Traverse}( \ec{t}, U, \emptyset ).$
\end{itemize}
\end{enumerate}

@@

\begin{enumerate}
\item If $\ec{t} \in P,$ $\rn{Acyclic}(t).$
\item If $\ec{t} \in U$:
\begin{enumerate}
\item[2.1.] If $\const{C}( t_1, \ldots, t_k ) \in \ec{t},$ then for each $j = 1, \ldots k$:
\begin{itemize}
\item[2.1.1.] $\rn{Traverse}( \ec{t_j}, U, P \cup \{ \ec{t} \} ).$
\end{itemize}
\item[2.2.] $U := U \textbackslash \{ \ec{t} \}.$
\end{enumerate}
\end{enumerate}
\vspace*{-\topsep}
\caption{\,Algorithm for applying the \rn{Acyclic} rule}
\label{fig:acyclic-algo}
\end{figure}

@@

\item Repeat until $\mathcalx{U}$ is unchanged:
\begin{enumerate}
\item[2.1.] $\mathcalx{U}' := \emptyset.$
\item[2.2.] For each $U_i \in \mathcalx{U}$:
\begin{itemize}
\item[2.2.1.] Let $\mathcalx{U}_{\vthinspace i}$ be a partition of $U_i$ such that $\ec{t} =_{\mathcalx{U}_{\vthinspace i}} \ec{u}$ for distinct $\ec{t}, \ec{u}$ if and only if
\begin{itemize}
\item $\const{C}( t_1, \ldots, t_k ) \in \ec{t}$ and $\const{C}( u_1, \ldots, u_k ) \in \ec{u},$ and
\item $\ec{t_j} =_{\mathcalx{U}} \ec{u_j}$ for each $j = 1, \ldots, k.$
\end{itemize}
\item[2.2.2.] $\mathcalx{U}' := \mathcalx{U}' \cup \mathcalx{U}_{\vthinspace i}.$
\end{itemize}
\item[2.3.] $\mathcalx{U} := \mathcalx{U}'.$
\end{enumerate}
\item If distinct $\ec{t}, \ec{u} \in U$ for some $U \in \mathcalx{U},$ $\rn{Unique}(t,u).$
\end{enumerate}
\vspace*{-\topsep}
\caption{\,Algorithm for applying the \rn{Unique} rule}
\label{fig:uniqueness-algo}
\end{figure}

@@

By contrast, the contributions of the decision procedure
to proving are less obvious; they depend on how
often acyclicity and uniqueness are necessary for a proof.
To evaluate this,
we generated benchmark problems from existing interactive
proof goals arising in existing
Isabelle formalizations, using Sledgehammer \cite{blanchette-et-al-2013-smt}
as translator from Isabelle to SMT-LIB.
%There are not yet many benchmarks available, since codatatypes are a
%recent addition to Isabelle \cite{blanchette-et-al-2014-impl}. \rem{Necessary to mention this? Seems to detract from significance of contribution.}
We
included all the formalizations from the Isabelle distribution (Distro, 1179~goals)
and the \emph{Archive of Formal Proofs} (AFP, 3014~goals) \cite{klein-et-al-afp}
that define codatatypes falling within the supported fragment. We
also included formalizations about Bird and Stern--Brocot trees (\gandl,
265~goals) \cite{gammie-lochbihler-2016}. To exercise the datatype support, formalizations about finite
lists and trees
were added to the first two benchmark sets. The formalizations were selected before
conducting any experiments. The experimental data are available
online.%
\footnote{\url{http://lara.epfl.ch/~reynolds/JAR-CADE2015-cdt/} or
\url{http://www21.in.tum.de/~blanchet/JAR-CADE2015-cdt/}}
